It all started with a phone call.\\
\par
At id software, we had enjoyed great success with our Commander Keen games, based on my childhood, my love of science fiction, and Chuck Jones' Warner Brothers cartoons. Now John Romero, one of the four co-founders of id (along with myself, John Carmack, and Adrian Carmack -- no relation) talked with Paul Neurath of Looking Glass Studios about a new game they were making (Ultima Underworld).  Romero got off the phone and said, "Paul said they are doing a thing called 'texture mapping'." John Carmack looked up and to his left, thought a bit, and said, "I can do that."\\
\par
We then started on a path to that goal. Hovertank 3D (also known as Hovertank One) was a simple drive-through-the-colored-wall maze shooter. Drive your tank with its turret in the bottom center of the screen. Shoot monsters, rescue humans. It, and the games that followed, each could be considered the first FPS (first-person shooter). Next was Catacomb 3D, a three-dimensional version of a 2D action game we made, but this time with textured walls (in 16-color EGA) and a hand casting magic fireballs.  Sometimes a fireball would hit a wall and it would disappear, revealing a secret room or path. We were then ready to try 256-color VGA.  We were looking for a theme. I came up with a hundred different ideas... but we needed something simple and brutal, like the engine was.\\
\par
Romero hit upon the idea of old Apple ][ game, Castle Wolfenstein. We'd all played it, and it seemed a natural fit. After some folly trying to replicate Beyond Castle Wolfenstein (the sequel to Castle Wolfenstein), which had body-dragging, disguises and such -- we settled on the simpler original. I came up with the base set of weapons, with the honor of figuring out what the first natural set of incrementally better FPS weapons should be. We were working on it for 14-18 hours, seven days a week. One night we looked at each other, saying, "No one's made a game like this before...." Sort of like, "Is that okay?" It was a weird, but exciting feeling. This 3D game was different.\\
\par
% Carmack had done something amazing and new with this engine.  We'd released two technical first-person shooters, each arguably the first free-moving 3D shooter... but for this real first-person shooter game we were a tad later than the first FULL 3D role-playing game that was our inspiration: Ultima Underworld, an amazing feat, was released on March 21, 1992.  Wolfenstein 3D, the first widely-known FPS, came out May 5th, 1992, coincidentally Adrian's birthday!\\
% \par
Commander Keen made good money. The next Keens made great money. Wolfenstein 3D made TEN TIMES that.  We were on to something.\\
\par
% But it all comes down to the technology. 
Carmack's brilliant mind had come up with this ray-casting lightning strike of an innovation, changing gaming forever. It's now quaint, but at the time, the height of gaming technology.  It takes a smart mind to explain its secrets and methods.\\
\par
Luckily, we have Fabien.  With great experience digging into classic games and the rendering engines behind them, there's no one better to show everyone the man behind the curtain. The Great and Powerful Oz seems magical, but is just great with math, geometry, algebra, and efficiency of algorithm. (And, well, not that great at spelling, heh, but that ain't important here.)\\
\par
We said we're here to see the wizard. Fabien saw our ruby slippers of intellectual curiosity.
Fabien lit up and said, "Why didn't you say that in the first place? That's a horse of a different color! Come on in!" And opened the gates.\\
\par
It's time to walk inside.\\
\par
-- Tom Hall\\
Game Designer;  Co-Founder and former Creative Director, id software\\
\par
San Francisco, CA\\
September, 2018\\
\thispagestyle{plain} % Trick to remove header